---
title: "Firm-Level Investment Responses to Oil Price Shocks"
subtitle: "An End-to-End Empirical Research Workflow in Python and R"
author: "Shuhan Zhang"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Project overview

This project examines whether oil-price shocks reduce firm-level capital
investment among U.S. manufacturing firms and whether the response is larger in
energy-intensive industries. It demonstrates an end-to-end empirical workflow:
Python is used to build the analytical firm-year panel, while R is used for
firm-specific fiscal-year alignment, panel construction, econometric estimation,
inference, and robustness analysis.

The workflow combines four data sources:

- Compustat firm-level financial statements;
- monthly West Texas Intermediate oil prices from FRED;
- four-digit NAICS energy-intensity measures constructed from the Annual Survey
  of Manufactures; and
- monthly structural oil shocks from Baumeister and Hamilton.

The proprietary Compustat observations are not included. The code nevertheless
uses relative paths, explicit input checks, and documented variable definitions
so that the analysis can be reproduced when authorized data are available.

# Analytical workflow

```text
Compustat data      \
WTI prices          ----> Python data preparation ----> Firm-year panel
ASM energy data     /                                      |
                                                           v
Structural shocks   ------------------------> R fiscal-year alignment
                                                           |
                         +---------------------------------+-------------------+
                         |                                 |                   |
                         v                                 v                   v
                 Reduced-form models              Dynamic effects      Shock decomposition
                         +---------------------------------+-------------------+
                                                           |
                                                           v
                                                Publication-ready outputs
```

# 1. Building the analytical dataset in Python

The companion script, `01_data_preparation.py`, performs the raw-data stage. It
filters the Compustat extract to manufacturing firms, constructs investment
outcomes and financial controls, handles gaps in the unbalanced panel, builds
annual WTI changes, constructs industry energy intensity, validates merge keys,
and exports the two processed files used in R.

A central data-quality rule is that a shifted value is treated as a one-year lag
only when the prior observation is from the immediately preceding fiscal year:

```python
def consecutive_lag(data, value_column, group_column="gvkey", year_column="fyear"):
    """Create a one-year lag without carrying values across panel gaps."""
    lagged_value = data.groupby(group_column)[value_column].shift(1)
    lagged_year = data.groupby(group_column)[year_column].shift(1)
    return lagged_value.where(data[year_column].eq(lagged_year + 1))
```

This prevents observations separated by multiple years from being incorrectly
interpreted as consecutive records.

# 2. R project setup and validation

The full statistical workflow is contained in `02_panel_analysis.R`. The script
first checks package availability, defines portable project paths, and validates
that required variables and firm-year keys are present before estimation.

```{r validation-example, eval=FALSE}
assert_unique_panel <- function(data, firm_id, year_id) {
  duplicated_key <- duplicated(data[c(firm_id, year_id)])

  if (any(duplicated_key)) {
    stop("The analytical panel contains duplicate firm-year observations.")
  }

  invisible(data)
}
```

These checks make assumptions about the analytical unit explicit and cause the
script to fail early when an input file is inconsistent with the research
design.

# 3. Constructing valid panel variables

Energy-intensive industries are defined using the 75th percentile of the
industry energy-intensity distribution. The preferred reduced-form exposure is
the interaction between a positive oil-price shock and this predetermined
industry indicator. Firm controls enter with a one-year lag to reduce mechanical
simultaneity with current investment.

```{r variable-construction, eval=FALSE}
energy_cutoff <- quantile(
  firm_panel$energy_intensity,
  probs = 0.75,
  na.rm = TRUE
)

firm_panel <- firm_panel |>
  mutate(
    high_energy = as.integer(energy_intensity >= energy_cutoff),
    oil_shock_positive = pmax(oil_shock, 0),
    oil_positive_high = oil_shock_positive * high_energy
  ) |>
  arrange(gvkey, fyear) |>
  group_by(gvkey) |>
  mutate(
    cash_at_w_l1 = consecutive_lag(cash_at_w, fyear),
    tangibility_w_l1 = consecutive_lag(tangibility_w, fyear)
  ) |>
  ungroup()
```

# 4. Aligning monthly shocks with firm-specific fiscal years

This is the most important data-engineering step in the R analysis. Firms have
heterogeneous fiscal year-end dates, so a calendar-year average can assign
monthly shocks to the wrong accounting period. The code constructs a 12-month
window for every firm-year and uses a `data.table` non-equijoin to match monthly
structural shocks to that window.

```{r fiscal-year-alignment, eval=FALSE}
firm_fiscal_windows <- firm_panel |>
  mutate(
    fiscal_end_month = floor_date(datadate, unit = "month"),
    fiscal_start_month = fiscal_end_month %m-% months(11)
  )

firm_dt <- as.data.table(firm_fiscal_windows)
shock_dt <- as.data.table(monthly_shocks)

matched_months <- shock_dt[
  firm_dt,
  on = .(
    date >= fiscal_start_month,
    date <= fiscal_end_month
  ),
  allow.cartesian = TRUE,
  nomatch = 0L
]
```

The matched monthly observations are then aggregated within each firm-year.
The script verifies the number of months matched before using the resulting
exposures in the regression analysis.

# 5. Main fixed-effects models

The preferred specifications include firm and fiscal-year fixed effects with
standard errors clustered by both firm and year. Firm fixed effects absorb
persistent differences across companies, while year fixed effects absorb common
macroeconomic shocks. Two-way clustering allows residual dependence within firms
over time and across firms exposed to the same annual environment.

```{r main-model, eval=FALSE}
model_dynamic <- feols(
  inv_at_w ~ oil_positive_high + oil_positive_high_l1 +
    ln_at_l1 + lev_w_l1 + prof_w_l1 + sales_growth_w_l1 +
    cash_at_w_l1 + tangibility_w_l1 |
    gvkey + fyear,
  data = dynamic_sample,
  vcov = ~ gvkey + fyear
)
```

The current-and-lagged model is estimated on a common complete-case sample so
that coefficient changes across specifications are not driven by changes in the
underlying observations.

# 6. Structural oil-shock decomposition

A change in the oil price can reflect different economic forces. The analysis
therefore separates supply-driven shocks from shocks associated with global
economic activity. Monthly structural shocks are aligned to firm fiscal years
using the same procedure described above, interacted with high energy intensity,
and entered in separate dynamic panel models.

```{r decomposition-model, eval=FALSE}
model_supply_dynamic <- feols(
  inv_at_w ~ supply_high_fy + supply_high_fy_l1 +
    ln_at_l1 + lev_w_l1 + prof_w_l1 + sales_growth_w_l1 +
    cash_at_w_l1 + tangibility_w_l1 |
    gvkey + fyear,
  data = decomposition_sample,
  vcov = ~ gvkey + fyear
)

model_activity_dynamic <- feols(
  inv_at_w ~ activity_high_fy + activity_high_fy_l1 +
    ln_at_l1 + lev_w_l1 + prof_w_l1 + sales_growth_w_l1 +
    cash_at_w_l1 + tangibility_w_l1 |
    gvkey + fyear,
  data = decomposition_sample,
  vcov = ~ gvkey + fyear
)
```

This step demonstrates how the empirical design moves beyond a single reduced-
form price measure and investigates the underlying source of observed oil-price
movements.

# 7. Cumulative dynamic effects

For models containing both current and lagged exposure, the economically
relevant effect may be the sum of the two coefficients. A reusable helper
computes the cumulative estimate and its standard error using the full
variance-covariance matrix, including the covariance between coefficients.

```{r cumulative-effect, eval=FALSE}
cumulative_effect <- function(model, current_term, lagged_term) {
  coefficients <- coef(model)
  variance <- vcov(model)

  estimate <- coefficients[current_term] + coefficients[lagged_term]
  standard_error <- sqrt(
    variance[current_term, current_term] +
      variance[lagged_term, lagged_term] +
      2 * variance[current_term, lagged_term]
  )

  tibble(
    estimate = estimate,
    standard_error = standard_error,
    statistic = estimate / standard_error
  )
}
```

# 8. Selected robustness checks

The portfolio version intentionally limits robustness analysis to two checks
that test distinct modeling choices.

## 8.1 Alternative investment outcome

The main outcome scales capital expenditures by beginning total assets. The
first robustness model instead scales capital expenditures by beginning net
property, plant, and equipment. This tests whether the result depends on the
normalization used to measure investment.

```{r alternative-outcome, eval=FALSE}
model_alternative_outcome <- feols(
  inv_ppent_w ~ oil_positive_high + oil_positive_high_l1 +
    ln_at_l1 + lev_w_l1 + prof_w_l1 + sales_growth_w_l1 +
    cash_at_w_l1 + tangibility_w_l1 |
    gvkey + fyear,
  data = alternative_outcome_sample,
  vcov = ~ gvkey + fyear
)
```

## 8.2 Continuous and nonlinear energy intensity

The second check replaces the high-energy indicator with continuous exposure
and a natural-spline specification. It tests whether the finding is an artifact
of imposing a 75th-percentile threshold and allows the response to vary
nonlinearly across the energy-intensity distribution.

```{r nonlinear-exposure, eval=FALSE}
model_energy_spline <- feols(
  inv_at_w ~ oil_spline_1 + oil_spline_2 + oil_spline_3 +
    ln_at_l1 + lev_w_l1 + prof_w_l1 + sales_growth_w_l1 +
    cash_at_w_l1 + tangibility_w_l1 |
    gvkey + fyear,
  data = functional_form_sample,
  vcov = ~ gvkey + fyear
)

wald(model_energy_spline, keep = "oil_spline")
```

Together, these checks address outcome measurement and functional-form
sensitivity without repeating a large number of closely related regressions.

# 9. Reproducible outputs

The R script exports labeled LaTeX tables, cumulative-effect summaries, and
session information. Keeping output generation in the analytical script reduces
manual transcription and ensures that tables remain synchronized with the
estimated models.

```{r output-example, eval=FALSE}
etable(
  list(
    "Reduced form" = model_dynamic,
    "Supply shock" = model_supply_dynamic,
    "Activity shock" = model_activity_dynamic
  ),
  tex = TRUE,
  file = here("output", "tables", "main_results.tex"),
  replace = TRUE
)

writeLines(capture.output(sessionInfo()), here("output", "sessionInfo.txt"))
```

# Files in this portfolio

- `README.md`: project overview, structure, and reproducibility notes;
- `01_data_preparation.py`: raw-data preparation and analytical panel creation;
- `02_panel_analysis.R`: complete documented R analysis;
- `02_panel_analysis.Rmd`: narrative portfolio notebook.

